---
title: "EPPS6326 & GISC6323: Unsupervised Machine-Learning with Geo-Referenced Data"
author: "Michael Tiefelsdorf & Yalin Yang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    df_print: paged
  pdf_document:
    toc: yes
vignette: |
 %\VignetteIndexEntry{DallasMarketAreas} 
 %\VignetteEncoding{UTF-8}
 %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r message=FALSE, warning=FALSE}
library(TexMix)    ## For mapping functions
library(sp);library(maptools)
library(spdep);library(ClustGeo)

#knitr::purl("E:\\Lectures2023\\GISC6323EPPS6326\\Labs\\Lab05\\SampleAnswerALab05MLSpring2023.Rmd", 
#             documentation = 0)

setwd("E:\\Lectures2023\\GISC6323EPPS6326\\Labs\\Lab05")

## Get neighboring state as spatial reference frame
neig.shp <- rgdal::readOGR(dsn="TXCnty2021", layer = "TXNeighbors",
                             integer64 = "allow.loss", stringsAsFactors=T)

## Get interstate layer for spatial orientation
interState.shp <- rgdal::readOGR(dsn="TXCnty2021", layer = "InterStateHwy",
                                 integer64 = "allow.loss", stringsAsFactors=T)

## Get polygons of TX counties
county.shp <- rgdal::readOGR(dsn="TXCnty2021", layer = "TXCnty2021",
                            integer64 = "allow.loss", stringsAsFactors=T)
```

# General modelling strategy

1.  Select those *features* which characterized the diversity among the
    spatial objects and that cover the objectives of the cluster
    analysis. The distribution of a selected variable should be
    outstandingly different at least for one cluster from the other
    clusters. See the section "Inspect feature profiles for each
    cluster". Are two variable profiles similar across all clusters then
    these variables are redundant with respect to the cluster analysis
    and one variable should be removed from the analysis.
2.  Select that *spatial dissimilarity metric* which leads to an
    interpretable regionalization without undermining the cluster
    homogeneities.
3.  First, identify the *number of clusters* $K$ while ignoring the
    regionalization, i.e., $\alpha=0$, by studying the dendogram and
    systematically increasing the number of clusters. Adopt that number
    of clusters which leads to interpretable results.
4.  Second, add *regionalization* to the optimal number of clusters by
    systematically increasing the mixture parameter $\alpha=0$. Select
    that solution, which reveals an interpretable regionalization and
    the profiles of clusters with respect to their characteristics. If
    necessary, some regions can be spatially fragmented.

Homogeneity within the clusters and their unique profiles are more
important than unfragmented regionalization.

# Prepare the feature similarity matrix

The selected features come from 5 logical groups of variables. These are
intended to describe the demographic, socio-economic, political,
cultural and infrastructure profile of each county.

**[a] cultural:** RELIGADHER and DISTMEX

**[b] political:** TRUMPVOT16 (%vote to Trump) and TURNOUT16

**[c] socio-economic:** UNINSURED, COLLEGEDEG, POVERTY and INCOME,

**[d] demographic:** MEDAGE, PARTBLACK, HISPORG and BIR15TO50

**[e] residential:** BNEW (Percent of housing units built after 2000),
POPDENSE, MEDVALHOME and POPCHG

Note: the population density `POPDENSE` should not be log-transformed.
Only untransformed will it be able to distinguish between rural,
sub-urban and metropolitan counties.

```{r}
## Exclude Loving County
county.shp <- county.shp[county.shp$NAME_x!="Loving", ]

Vars <- county.shp@data
Vars <- transform(Vars,
  RELIGADHER = as.numeric(RELIGADHER),
  TRUMPVOT16 = TRUMPVOT16/TOTALVOT16,
  BNEW = B2000PCT+B2010PCT
)

varKeep <- c("RELIGADHER","DISTMEX", "TRUMPVOT16", "TURNOUT16",
             "UNINSURED", "COLLEGEDEG", "POVERTY", "INCOME",
             "MEDAGE", "PARTBLACK", "HISPORG", "BIR15TO50",
             "POPDENSE", "POPCHG", "BNEW", "MEDVALHOME")

xVars <- Vars[, varKeep]
row.names(xVars) <- 1:nrow(xVars)
featDist <- dist(scale(xVars))
```

# Definition of the spatial dissimilarity metric

This section extracts from the counties their underlying dissimilarity
metrics. The three competing metrics are:

-   Basic spatial adjacency among the counties
-   Spherical distances among the counties
-   Shortest path distances among the counties

## Spatial topology

```{r fig.height=7, fig.width=7}
county.bbox <- bbox(county.shp)                # county bounding box for map region
county.centroid <- coordinates(county.shp)     # Get county centroids

##
## Get spatial structure distance matrices
##
nb <- spdep::poly2nb(county.shp, queen=F)      # extract first order neighbors links
B <- spdep::nb2mat(nb, style="B")              # convert neighbor list to binary matrix
plot(county.shp, col="palegreen3", border=grey(0.9), axes=T) # map topology
plot(nb, coords=coordinates(county.shp), pch=19, cex=0.1, col="blue", add=T)
title("Spatial Neighbors Links among TX County") 
topoDist <- 1-B; diag(topoDist) <- 0          # convert into dissimilarity
topoDist <- as.dist(topoDist)                 # convert into distance object
```

## Shortest path distance

```{r}
##
## Get steps distance from topology
##
BNa <- ifelse(B==0,NA,B)                      # recode 0's to NA's
allPath <- e1071::allShortestPaths(BNa)       # calculate the shortest path among all nodes
pathDist <- as.dist(allPath$length)           # number of steps from origin to destination node
```

## Spherical distance

```{r}
sphDist <- sp::spDists(county.shp, longlat=T)   # spherical distance matrix among tracts in km
sphDist <- as.dist(sphDist)
```

In this study the spherical distance among the counties, i.e.,
`sphDist`, lead to the most interpretable results. The clusters became
mostly spatially connected based on their topological relationship.
Experimentation with other spatial distance matrices, i.e., `pathDist`
or `topoDist`, lead to less desirable outcomes with higher regional
fragmentation and feature heterogeneity within the derived clusters.

# Evaluate mixture of feature and spatial dissimilarity

Through experimentation, setting the number of clusters to $K=7$ gives
reasonable and interpretable results with respect to both the
homogeneity within clusters and visual interpretation. The clusters
mostly were already spatially grouped. In addition, setting the spatial
mixture parameter $\alpha=0.1$ maintained the cluster homogeneity while
enhancing the regionalization.

```{r fig.height=9, fig.width=9}
K <- 7                             # Number of distinct clusters
range.alpha <- seq(0, 1, by=0.1)   # Evaluation range
cr <- choicealpha(featDist, sphDist, range.alpha, K, graph=TRUE)
tree <- hclustgeo(featDist, sphDist, alpha=0.1)
plot(tree, hang=-1)
rect.hclust(tree, k=K)
```

```{r ig.height=7, fig.width=9}
neighClus <- as.factor(cutree(tree, K))        # Determine cluster membership
## Label clusters
neighClus <- factor(neighClus, labels=c("Historic Black", "Religious", "Affluent Rural", 
                                        "Conservative", "Hispanic", "Metropolitan", "Suburban") )
table(neighClus)
plot(table(neighClus))                         # number of counties in each cluster
```

# Map Results

```{r fig.height=9, fig.width=9}
plot(neig.shp, axes=T, col=grey(0.9),                  # first background (only for final maps)
     border="white", bg="lightblue",
     xlim=county.bbox[1,], ylim=county.bbox[2,])        

mapColorQual(neighClus, county.shp, 
             map.title ="Spatially Constrained Cluster Analysis",
             legend.title="Cluster\nId.", legend.cex=0.9, add=T)

plot(interState.shp, col="tomato4", lwd=1, add=T)      # insert road network for orientation

```

# Cluster interpretation

Each cluster has a unique feature profile and none of the feature
distributions for the clusters are identical.

## Feature profiles for each cluster

```{r fig.height=12, fig.width=9}
plotBoxesByFactor(xVars[,1:4], neighClus, ncol=1, zTrans=T, varwidth=F)
plotBoxesByFactor(xVars[,5:8], neighClus, ncol=1, zTrans=T, varwidth=F)
plotBoxesByFactor(xVars[,9:12], neighClus, ncol=1, zTrans=T, varwidth=F)
plotBoxesByFactor(xVars[,13:16], neighClus, ncol=1, zTrans=T, varwidth=F)
```

## Numeric cluster description

```{r }
options(digits=2)
by(scale(xVars), neighClus, colMeans)
```

## Heatmap of features and counties

```{r fig.width=9, fig.height=9}
library("RColorBrewer")
col <- rev(colorRampPalette(brewer.pal(10, "RdYlBu"))(64))

## Calculate hclust dendrograms
hclust_rows <- as.dendrogram(tree)  
hclust_cols <- as.dendrogram(hclust(dist(t(scale(xVars)))))

## Draw heatmap with hclust dendrograms
heatmap(as.matrix(scale(xVars)), col=col,                                
        Rowv = hclust_rows, Colv = hclust_cols)
```

## Cluster and region descriptions

-   **Hispanic**: This cluster is located predominately along the Rio
    Grande Valley and the Western Mexican border with the highest
    concentration of the Hispanic population. This cluster is quite
    distinct: lower republican votes and election participation, higher
    uninsured and poverty rates, a younger population with a higher
    birth rate and otherwise being economically depressed.
-   **Affluent Rural**: This cluster is overall around average for rural
    counties. However, it is above average for many economic measures.
    Counties in the cluster are predominately close to urban centers. In
    West Texas counties in the cluster are close to predominately white
    religious and republican counties.
-   **Religious**: Overall the religious affiliation in this cluster is
    high which is followed by a high republican orientation. Otherwise,
    counties in this cluster are about average with a slightly lower
    African-American population. This cluster stretches all the way down
    from the Pan-Handle to the Mexican border.
-   **Conservative**: Similar to the *religious* cluster with the
    difference that the republican affiliation is even higher, a lower
    Hispanic and African-American population, an older population and
    overall being economically better situated. Counties in this cluster
    don't tend to form regions and they are mostly embedded in the
    *religious* cluster.
-   **Suburban**: Counties in this cluster are adjacent to
    *metropolitan* counties, but with a lower population density than
    the *metropolitan* counties. There is a below average religious and
    Republican affiliation, highest economic and education status, lower
    birth rates but younger population, and newer, more expensive homes.
    Its population is on average increasing due to immigration.
-   **Historic Black**: Counties in the rural cluster have a higher
    fraction of African-Americans but lower Hispanics population. The
    educational attainment is below average. The republican affiliation
    is slightly above averages. Counties in this cluster are
    concentrated in East Texas. The higher numbers of African-American
    population has historic roots.
-   **Metropolitan**: The 4 counties in this cluster are Dallas and Fort
    Worth, Houston and San Antonio. These counties have the highest
    population density. Its population on average is younger and
    increasing due to immigration. Homes are newer and more expensive. A
    larger proportion of African-Americans and Hispanics are living in
    these counties, which makes the economic profile mixed, but the
    educational attainment is higher. The political orientation of
    population in these metropolitan counties is more democratically
    oriented. Interestingly, ElPaso is not included in this cluster due
    to its specific Hispanic profile.
